---
title: "Gun Violence Data EDA"
author: "Markus Mayer"
date: "6 Juli 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Plotting
# install.packages('ggplot2', dep = TRUE)
library(ggplot2)
library(gridExtra)

# lubridate for date processing
# install.packages('lubridate')
library(lubridate)

# tidyr and dplyr for data wrangling
# install.packages('tidyr')
# install.packages('dplyr')
library(tidyr)
library(dplyr)
```

## Data source

The data inspected in this project is taken from the [gun violence data](https://github.com/jamesqo/gun-violence-data) set:

> A comprehensive, accessible database that contains records of over 260k US gun violence incidents from January 2013 to March 2018.

## Structure of the dataset

```{r gv}
gv <- read.csv('dataset/stage3.csv', nrows=100)
```

The dataset consists of about 240k observations of 29 variables. In order to
show an issue with the dataset, only `100` rows were loaded:

```{r dimensions, echo=FALSE}
dim(gv)
```

The variables include the date and location of the incident, the number
of guns involved, as well as information such as age, gender and relationship
of the participants.

```{r names, echo=FALSE}
names(gv)
```

Here's the structure of the dataset after importing:

```{r gv structure, echo=FALSE}
str(gv)
```

Some variables such as `participant_gender` incorrectly appear as factors with
high number of levels.

```{r gender-factors, echo=FALSE}
str(gv$participant_gender)
```

This is due to the way data is encoded as lists; in the gender example, one
gender value is given for every person involved in the incident. Before
investigating the data further, these lists should be unraveled into rows.

In order to bring the data into a more usable format, the original CSV file
was preprocessed in the Jupyter Notebook
[dataset/split-dataset.ipynb](dataset/split-dataset.ipynb) using Python 
and Pandas, creating four separate files,

- `dataset/incidents.csv`,
- `dataset/participants.csv`,
- `dataset/gun_use.csv` and
- `dataset/characteristics.csv`

```{r}
rm(gv)
inc <- read.csv('dataset/incidents.csv')
par <- read.csv('dataset/participants.csv')
gun <- read.csv('dataset/gun_use.csv')
cha <- read.csv('dataset/characteristics.csv')
```

We now find that we have `239677` incident records of `15` variables each:

```{r echo=FALSE}
inc <- mutate(inc, 
       date = ymd(date), 
       address = as.character(address),
       location_description = as.character(location_description),
       congressional_district = as.factor(congressional_district),
       state_house_district = as.factor(state_house_district),
       state_senate_district = as.factor(state_senate_district),
       n_guns_involved = as.integer(n_guns_involved))
str(inc)
```

There are `392323` participant records of `10` variables each:

```{r echo=FALSE}
par <- mutate(par, participant_name = as.character(participant_name))
str(par)
```

There are `192452` gun use records of `4` variables each (excluding relations):

```{r echo=FALSE}
str(gun)
```

And finally

There are `579820` characteristics of `2` variables (excluding relations):

```{r echo=FALSE}
str(cha)
```

## Univariate Plots

First we're goint to look at the accumulated number of incidents over time.
The dataset description mentions that for the year 2013, not much data was
collected. We can see this as a highly noticeable dent at the January 1st,
2014.

```{r echo=FALSE}
# For each occurrence, we accumulate the total number of occurrences
# up to that date. We then group by date in order to obtain the number
# of incidents recorded up to that day.
occurrences <- aggregate(count ~ date, 
                     data = data.frame(
                                count=cumsum(rep(1, nrow(inc))),
                                date=inc$date), 
                     max)

ggplot(occurrences, aes(date)) +
    geom_line(aes(y=count)) +
    scale_x_date(date_breaks = "1 year", 
                 date_labels = '%Y', 
                 date_minor_breaks = "1 months") +
    ylab('occurrences (running total)') + 
    xlab('date')
```

Removing the year 2013 from the plot, we can compare this to the regression
line, a line for one incident per day and a line for 100 incidents per day.
We find that the actual numbers slowly deviate from the trendline in a 
quasi-cyclic manner; putting the numbers into relation with the hypothetical
"one incident per day" and "one hundred incidents per day" lines demonstrate
the sheer amount of incidents, being much higher than both of them.

```{r echo=FALSE, warning=FALSE}
occurrences_since_2014 <- subset(occurrences, date > date('2014-01-01'))

diff_since_2014 <- occurrences_since_2014
diff_since_2014$count <- occurrences_since_2014$date - 
                                    min(occurrences_since_2014$date)

diff100_since_2014 <- diff_since_2014 
diff100_since_2014$count <- diff_since_2014$count * 100

fit <- lm(formula = count ~ date,
          data = subset(occurrences_since_2014))

fit_since_2014 <- diff_since_2014 
fit_since_2014$count <- predict(fit, fit_since_2014)

occurrences_since_2014$type <- 'baseline'
diff_since_2014$type <- 'diff'
diff100_since_2014$type <- 'diff 100'
fit_since_2014$type <- 'fit'

df <- rbind(occurrences_since_2014, diff_since_2014, 
            diff100_since_2014, fit_since_2014)
rm(occurrences_since_2014, diff_since_2014, diff100_since_2014,  fit_since_2014)

labels <- c("actual numbers", "one per day",  "100 per day", "regression")

ggplot(df, aes(date)) +
    geom_line(aes(y=count, col=type, linetype=type)) +
    scale_y_continuous(limits = c(0, max(df$count))) +
    scale_x_date(date_breaks = "1 year", 
                 date_labels = '%Y', 
                 date_minor_breaks = "3 months") +
    scale_linetype_manual(labels = labels,
                          values = c("solid", "dashed", "twodash", "dotted")) +
    scale_color_manual(labels = labels, 
                       values = c("black", "darkgray", "darkred", "#333333")) +
    labs(title = "", # "Gun violence incidents since 2014", 
         x = "date", 
         y = "running total", 
         col = "trend line", linetype = "trend line") +
  theme(legend.position="bottom")
```

The regression line shows an average of `156.1` incidents per day.

```{r echo=FALSE}
fit$coefficients
```

When the (artificial) trend is removed, we obtain the number of
incidents relative to the expected total number of incidents. A value of
about 6000 for January 1st, 2014 implies that about 6000 more gun violence
incedents occurred than the trendline suggests. The smallest relative number
of incidents can be found at June 18th, 2016, with abour 3723 incidents
less than expected.

```{r echo=FALSE, warning=FALSE}
baseline <- subset(df, type == 'baseline')
regression <- subset(df, type == 'fit')
delta <- baseline
delta$count <- baseline$count - regression$count

delta[which.min(delta$count), ]
```

When the relative number of incidents is plotted, it becomes obvious
that this day also marks the turning point of incident numbers:
Starting this day, numbers begin to rise again.

```{r echo=FALSE, warning=FALSE}
delta$dir <- c(NA, diff(delta$count, lag = 1, differences = 1))
#delta$weight <- (delta$dir - min(delta$dir)) / 
#                (max(delta$dir) - min(delta$dir))
# delta$color <- rgb(colorRamp(c("red", "green"))(delta$weight) / 255)

a <- ggplot(delta, aes(date)) +
    geom_line(aes(y=count)) +
    scale_x_date(date_breaks = "1 year", 
                 date_labels = '%Y', 
                 date_minor_breaks = "3 months") +
    labs(title = "indicent counts relative to expectation",
         x = "date", 
         y = "# incidents relative to trend")

b <- ggplot(delta, aes(date)) +
    geom_smooth(aes(y=dir), method = 'loess', span = 0.1, 
                color = 'black', size = 0.5) +
    scale_x_date(date_breaks = "1 year", 
                 date_labels = '%Y', 
                 date_minor_breaks = "3 months") +
    labs(title = "approximate trend direction",
         x = "date", 
         y = "1st order trend difference")

grid.arrange(a, b, ncol=1)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
