---
title: "Gun Violence Data EDA"
author: "Markus Mayer"
date: "6 Juli 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Plotting
# install.packages('ggplot2', dep = TRUE)
library(ggplot2)
library(gridExtra)

# lubridate for date processing
# install.packages('lubridate')
library(lubridate)

# tidyr and dplyr for data wrangling
# install.packages('tidyr')
# install.packages('dplyr')
library(tidyr)
library(dplyr)

# rworldmap for plotting the incident locations
# install.packages('rworldmap')
library(rworldmap)

# percentages on plot scales
require(scales)
```

## Data source

The data inspected in this project is taken from the [gun violence data](https://github.com/jamesqo/gun-violence-data) set:

> A comprehensive, accessible database that contains records of over 260k US gun violence incidents from January 2013 to March 2018.

## Structure of the dataset

```{r gv}
gv <- read.csv('dataset/stage3.csv', nrows=100)
```

The dataset consists of about 240k observations of 29 variables. In order to
show an issue with the dataset, only `100` rows were loaded:

```{r dimensions, echo=FALSE}
dim(gv)
```

The variables include the date and location of the incident, the number
of guns involved, as well as information such as age, gender and relationship
of the participants.

```{r names, echo=FALSE}
names(gv)
```

Here's the structure of the dataset after importing:

```{r gv structure, echo=FALSE}
str(gv)
```

Some variables such as `participant_gender` incorrectly appear as factors with
high number of levels.

```{r gender-factors, echo=FALSE}
str(gv$participant_gender)
```

This is due to the way data is encoded as lists; in the gender example, one
gender value is given for every person involved in the incident. Before
investigating the data further, these lists should be unraveled into rows.

In order to bring the data into a more usable format, the original CSV file
was preprocessed in the Jupyter Notebook
[dataset/split-dataset.ipynb](dataset/split-dataset.ipynb) using Python 
and Pandas, creating four separate files,

- `dataset/incidents.csv`,
- `dataset/participants.csv`,
- `dataset/gun_use.csv` and
- `dataset/characteristics.csv`

```{r load, echo=FALSE}
rm(gv)
inc <- read.csv('dataset/incidents.csv')
par <- read.csv('dataset/participants.csv')
gun <- read.csv('dataset/gun_use.csv')
cha <- read.csv('dataset/characteristics.csv')
```

We now find that we have `239677` incident records of `15` variables each:

```{r mutate-inc, echo=FALSE}
inc <- mutate(inc, 
       date = ymd(date), 
       congressional_district = as.factor(congressional_district),
       state_house_district = as.factor(state_house_district),
       state_senate_district = as.factor(state_senate_district),
       n_guns_involved = as.integer(n_guns_involved),
       source_url = as.character(source_url),
       incident_url = as.character(incident_url))
str(inc)
```

There are `392323` participant records of `10` variables each:

```{r mutate-par, echo=FALSE}
par <- mutate(par, participant_name = as.character(participant_name))
str(par)
```

There are `192452` gun use records of `5` variables each (including relations):

```{r gun, echo=FALSE}
str(gun)
```

And finally

There are `579820` characteristics of `3` variables (including relations):

```{r cha, echo=FALSE}
str(cha)
```

## Univariate Plots

First we're goint to look at the accumulated number of incidents over time.
The dataset description mentions that for the year 2013, not much data was
collected. We can see this as a highly noticeable dent at the January 1st,
2014.

```{r accumulated-totals, echo=FALSE}
# For each occurrence, we accumulate the total number of occurrences
# up to that date. We then group by date in order to obtain the number
# of incidents recorded up to that day.
occurrences <- aggregate(count ~ date, 
                     data = data.frame(
                                count=cumsum(rep(1, nrow(inc))),
                                date=inc$date), 
                     max)

ggplot(occurrences, aes(date)) +
    geom_line(aes(y=count)) +
    scale_x_date(date_breaks = "1 year", 
                 date_labels = '%Y', 
                 date_minor_breaks = "1 months") +
    ylab('occurrences (running total)') + 
    xlab('date')
```

Removing the year 2013 from the plot, we can compare this to the regression
line, a line for one incident per day and a line for 100 incidents per day.
We find that the actual numbers slowly deviate from the trendline in a 
quasi-cyclic manner; putting the numbers into relation with the hypothetical
"one incident per day" and "one hundred incidents per day" lines demonstrate
the sheer amount of incidents, being much higher than both of them.

```{r totals-since-2014, echo=FALSE, warning=FALSE}
occurrences_since_2014 <- subset(occurrences, date > date('2014-01-01'))

diff_since_2014 <- occurrences_since_2014
diff_since_2014$count <- occurrences_since_2014$date - 
                                    min(occurrences_since_2014$date)

diff100_since_2014 <- diff_since_2014 
diff100_since_2014$count <- diff_since_2014$count * 100

fit <- lm(formula = count ~ date,
          data = subset(occurrences_since_2014))

fit_since_2014 <- diff_since_2014 
fit_since_2014$count <- predict(fit, fit_since_2014)

occurrences_since_2014$type <- 'baseline'
diff_since_2014$type <- 'diff'
diff100_since_2014$type <- 'diff 100'
fit_since_2014$type <- 'fit'

df <- rbind(occurrences_since_2014, diff_since_2014, 
            diff100_since_2014, fit_since_2014)
rm(occurrences_since_2014, diff_since_2014, diff100_since_2014,  fit_since_2014)

labels <- c("actual numbers", "one per day",  "100 per day", "trend line")

ggplot(df, aes(date)) +
    geom_line(aes(y=count, col=type, linetype=type)) +
    scale_y_continuous(limits = c(0, max(df$count))) +
    scale_x_date(date_breaks = "1 year", 
                 date_labels = '%Y', 
                 date_minor_breaks = "3 months") +
    scale_linetype_manual(labels = labels,
                          values = c("solid", "dashed", "twodash", "dotted")) +
    scale_color_manual(labels = labels, 
                       values = c("black", "darkgray", "darkred", "#333333")) +
    labs(title = "", # "Gun violence incidents since 2014", 
         x = "date", 
         y = "running total", 
         col = "trend line", linetype = "trend line") +
  theme(legend.position="bottom")
```

The regression line shows an average of `156.1` incidents per day.

```{r totals-regression-coeffs, echo=FALSE}
fit$coefficients
```

When the (artificial) trend is removed, we obtain the number of
incidents relative to the expected total number of incidents. A value of
about 5719 for January 2nd, 2014 implies that about 5719 more gun violence
incidents occurred than the trendline suggests.

```{r highest-number-rel-trend, echo=FALSE, warning=FALSE}
baseline <- subset(df, type == 'baseline')
regression <- subset(df, type == 'fit')
delta <- baseline
delta$count <- baseline$count - regression$count

max_rel_incidents_date = delta[which.max(delta$count), ]
max_rel_incidents_date
```

The smallest relative number
of incidents can be found at June 18th, 2016, with about 3723 incidents
less than expected.

```{r lowest-number-rel-trend, echo=FALSE, warning=FALSE}
min_rel_incidents_date = delta[which.min(delta$count), ]
min_rel_incidents_date
```

When the relative number of incidents is plotted, it becomes obvious
that this day also marks the turning point of incident numbers:
Starting this day, numbers begin to rise again. If the first-order differences
are plotted, this day marks the zero-crossing into the realm of positive
numbers, indicating change of curvature.

```{r change-in-totals, echo=FALSE, warning=FALSE}
delta$dir <- c(NA, diff(delta$count, lag = 1, differences = 1))
#delta$weight <- (delta$dir - min(delta$dir)) / 
#                (max(delta$dir) - min(delta$dir))
# delta$color <- rgb(colorRamp(c("red", "green"))(delta$weight) / 255)

a <- ggplot(delta, aes(date)) +
    geom_line(aes(y=count)) +
    geom_vline(xintercept=as.numeric(min_rel_incidents_date$date),
                col = "darkred", 
                lty = "dotted" ) +
    scale_x_date(date_breaks = "1 year", 
                 date_labels = '%Y', 
                 date_minor_breaks = "3 months") +
    labs(title = "indicent counts relative to expectation",
         x = "date", 
         y = "# incidents relative to trend")

b <- ggplot(delta, aes(date)) +
    geom_smooth(aes(y=dir), method = 'loess', span = 0.1, 
                color = 'black', size = 0.5) +
    geom_vline(xintercept=as.numeric(min_rel_incidents_date$date),
                col = "darkred", 
                lty = "dotted" ) +
    scale_x_date(date_breaks = "1 year", 
                 date_labels = '%Y', 
                 date_minor_breaks = "3 months") +
    labs(title = "approximate trend direction",
         x = "date", 
         y = "change in incidents/day rel. to trend")

grid.arrange(a, b, ncol=1)
```

Next is the distribution of incident counts per state. 
The numbers differ by a fair amount, but are also not taking the size of the
state's population into account, i.e. the number of incidents per capita
could paint a different picture.
Since the state, city, address, district, latitude and longitude information is
correlated, we would expect to find similar information in these plots as well.
The plot also shows the national mean (dotted) and median (dashed, gray)
taken over the whole time range.

```{r incidents-per-state, echo=FALSE, fig.width=6, fig.height=8}
national_mean <- mean(summary(inc$state))
national_median <- median(summary(inc$state))

ggplot(aes(state), data=inc) +
    geom_bar(width = 0.5) +
    coord_flip() +
    scale_x_discrete(limits = rev(levels(inc$state))) +
    geom_abline(slope=0, intercept=national_mean,  
                col = "darkred", 
                lty = "dotted" ) +
    geom_abline(slope=0, intercept=national_median, 
                col = "black", 
                lty = "dashed",
                alpha = 0.5)
```

The next look would be at the cities with the most recorded incidents (limited
to 20 for brevity). We find that the city with the most recorded incidents
is Chicago, IL, which coincides with the highest number of recorded incidents
in any state. However, the second city in the list, Baltimore, is in Maryland,
which has less than half the number of incidents of Illnois.
Again, the numbers are put in relation to national mean (dotted red) and
median (dashed gray) over the whole time range.

```{r incidents-per-city, echo=FALSE}
# Most ocurrences are "other", so we skip that
cities_most <- sort(summary(inc$city_or_county), decreasing=T)[2:21]
cities_most <- data.frame(name = factor(names(cities_most), 
                                        levels = names(cities_most)),
                          count = as.integer(cities_most),
                          stringsAsFactors = T)
ggplot(cities_most) +
    geom_bar(aes(name, y = count), stat = "identity") +
    coord_flip() +
    scale_x_discrete(limits = rev(levels(cities_most$name))) +
    geom_abline(slope=0, intercept=national_mean,  
                col = "darkred", 
                lty = "dotted" ) +
    geom_abline(slope=0, intercept=national_median, 
                col = "black", 
                lty = "dashed",
                alpha = 0.5)

rm(cities_most)
```

Interestingly, we do find that Illinois has three times the number of incidents
as Maryland, while Chicago has 2.7 times the number of incidents as Baltimore.

```{r, echo=FALSE, include=FALSE}
sum(inc$state == "Illinois") / sum(inc$state == "Maryland")
sum(inc$city_or_county == "Chicago") / sum(inc$city_or_county == "Baltimore")
```

The Wikipedia article [Crime in Chicago](https://en.wikipedia.org/wiki/Crime_in_Chicago)
reports the violent crime rate of Chicago as being higher than the U.S. average,
while also being responsible for half of the increase in homicides in the
year 2016. Likewise, the article [Crime in Baltmore](https://en.wikipedia.org/wiki/Crime_in_Baltimore)
ranks the cities' violent crime rate "high above the national average".
In the above plot we find Baltimore below the national average, however the
dataset only covers gun violence specifically; here, we do indeed find
Baltimore above the median.

The next thing to look at would be the number of guns involved in the incidents.
We find that significantly more incidents happen with a small number of guns,
but also that the distribution is somewhat long-tailed. High number of guns
are apparently reported in distinct steps, i.e. there's some occurrences of 100, 200,
300 and 400 guns reported.

```{r guns-counted-log10, echo=FALSE, warning=FALSE}
guns_counted <- subset(inc, !is.na(inc$n_guns_involved) & 
                           inc$n_guns_involved > 0)
ggplot(guns_counted) +
    # geom_histogram(aes(n_guns_involved), binwidth=1) +
    geom_point(aes(n_guns_involved), stat = "count", alpha=1.0) +
    scale_y_log10()
```

What is obvious is that the points appearing at multiples of ten stand out
from the remaining values by following a different curve.
While incidents with more than 50 guns appear at most once in general,
events with e.g. 100, 200, 300 and 400 are reported significantly more often.
Given that there are entries for 399 guns (a single time) and 400 guns 
(four times), it could be that entries are duplicated. This turns out to be not
the case, with entries being at different dates and states.

```{r, echo=FALSE}
subset(inc, inc$n_guns_involved > 350)[, c('date', 'state', 'n_guns_involved')]
```

When log-log transforming the data, we find that the values appear to 
approximately approximately follow a log-log linear distribution with the
probability of a given number of incidents being inversely proportional to the 
number of guns involved in the crime (much like a
[Zipfian distribution](https://en.wikipedia.org/wiki/Zipf%27s_law)); we 
also see that the plot exhibits heteroscedastic behavior (that is, variance of 
occurrences grows with the number of guns involved). After an amount of about 
100 guns per incident, values appear to be erratic under this assumption.

```{r guns-counted-loglog, echo=FALSE, warning=FALSE}
ggplot() +
    geom_point(aes(n_guns_involved), data = guns_counted, stat = "count") +
    geom_line(aes(y = c(2e4, 1), x=c(1, 90)), 
              linetype = "dotted", alpha = 0.5, color = "darkred") +
    scale_y_log10() +
    scale_x_log10()
```
Likewise, we might want to know how many people were involved in each incident:

```{r participants-log10, echo=FALSE, warning=FALSE}
participants_counted <- subset(inc, !is.na(inc$incident_participants))
ggplot(participants_counted) +
    geom_point(aes(incident_participants), stat = "count", alpha=1.0)
```

We find that we have an extremely high number of incidents with exactly one
person and logarithmically less incidents with more than one person. We also
observe two interesting situations: There seems to be a single incident with 100 
people involved, as well as about 25k incidents with no people involved at all.

When we investigate the issue with the most participants, we find it to be the 
[Orlando nightclub shooting](https://en.wikipedia.org/wiki/Orlando_nightclub_shooting) 
at December 12th, 2016, where a 29-year old security guard killed 49 nightclub
visitors in a terroristic attack, injured 50 more and eventually got 
killed by the police.

```{r max-incidents, echo=FALSE, warning=FALSE}
inc[which.max(inc$incident_participants), ]
```

There are some more variables we can look at; here's the distribution of gun
types for all incidents where it is known:

```{r gun-type, echo=FALSE}
ggplot(aes(gun_type), data=gun[gun$gun_type != 'Unknown', ]) +
    geom_bar(width = 0.5) +
    coord_flip()
```

Next is the distribution of age ranges; note that this statistic includes the 
ages of all participants, i. e. includes victims.

```{r, age-groups, echo=FALSE}
par$participant_age_group_ordered <- factor(par$participant_age_group, 
                                            levels=c('Child 0-11', 'Teen 12-17', 
                                                     'Adult 18+', ''), ordered=TRUE)
ggplot(aes(participant_age_group_ordered), data=par) +
    geom_bar(aes(y = (..count..)/sum(..count..)), width = 0.5) +
    scale_y_continuous(labels=percent)
```

Apart from the unknown block, there is a single occurrence of a `Male, female`
case.

```{r, male-female, echo=FALSE}
par[par$participant_gender == 'Male, female', ]
```

When investigating the case, we find that it is about a [drive-by shooting in
Marquette Park, Chicago](https://www.dnainfo.com/chicago/20141211/marquette-park/marquette-park-drive-by-wounds-woman/).
Here, a 32 year old woman was wounded by an unknown perperatror. It seems that
the label is single wrong.

```{r, male-female-incident, echo=FALSE}
inc[inc$incident_id == 261784, c('incident_id', 'date', 'source_url')]
```

We can correct this issue by assigning the correct gender to the case.

```{r, include=FALSE}
par[par$participant_gender == 'Male, female', ]$participant_gender = 'Female'
```

When we look at the actual ages, we find that the median is 26 years, and that
75% of all participants were younger than 35 years, however there appear to be
some outliers on the lower and upper ends:

```{r, age-stats, echo=FALSE}
summary(par$participant_age)
```

Let's just quickly see cases where a participant was reported as older than
100 years, an arbitrarily chosen threshold.

```{r, age-list, echo=FALSE}
par_extreme_ages <- par[!is.na(par$participant_age) & par$participant_age > 100, ]
par_extreme_ages
```

Looking at the source material for these incidents it turns out that the cases
with 101 year old participants are actually correct, whereas the bigger numbers
are wrong. We can thus fix them.

```{r, age-list-extreme, echo=FALSE}
inc_par <- right_join(inc, par_extreme_ages, by = 'incident_id')
inc_par[, c('incident_id', 'date', 'participant_age', 'incident_url')]
```

```{r, fix-ages, include=FALSE}
par[par$incident_id == 926498, ]$participant_age <- 33
par[par$incident_id == 1001184, ]$participant_age <- 31
```

We can now look at the participant age distribution.

```{r, ages, echo=FALSE}
ggplot(aes(participant_age), data=par[!is.na(par$participant_age), ]) +
    geom_bar(aes(y = (..count..)/sum(..count..)), width = 0.5) +
    scale_y_continuous(labels=percent)
```

When we next we look at the distribution of participant gender, we find that
the participants are mostly male and the number of male participants is
eightfold the number of female and unknown ones.

```{r, gender, echo=FALSE}
num_items <- as.integer(count(par))
ggplot(aes(participant_gender), data=par) + 
  geom_bar(aes(y = (..count..)/sum(..count..)), width = 0.5) +
  scale_y_continuous(labels=percent, name = 'share of participants', 
                     sec.axis = sec_axis(~ . * num_items / 1000, name = "number of participants (in thousands)"))
```



## Bivariate Plots

The following map shows the locations of incidents. Each incident is marked
with 1% transparency, i.e. it takes (at least) one hundred incidents in the very
same location to render the marker fully red. As we can see on the map,
locations of high incident counts seemto cluster around the western coast of the
USA and fill out the entire eastern half of the USA; at the same time, Hawaii
also pops up in red, with two visible peaks indicating at least 200 incidents.

```{r map-of-incidents, echo=FALSE, warning=FALSE}
worldmap <- getMap(resolution = 'low')
usa <- worldmap[which(grepl('US', worldmap$GEO3)), ]
usa$id_1 <- row.names(usa)

map <- fortify(usa)
valid_locations = !is.na(inc$longitude) & !is.na(inc$latitude)
incident_with_coords <- subset(inc, valid_locations)

ggplot() + 
  coord_map(xlim = c(-180, -60), ylim = c(10, 70)) +
  geom_polygon(data = map, aes(x = long, y = lat, group = group), fill = 'gray22', alpha = 1) +
  geom_point(data = incident_with_coords, aes(x = longitude, y = latitude), color = "red", alpha = 0.01)
```

When we count all incidents in Hawaii only, we find that we indeed have 289 occurrences:

```{r hawaii-count, echo=FALSE}
count(inc[inc$state == 'Hawaii', ])
```

The effect of having a high incident counts in eastern USA in contrast to a
comparatively lower incident counts in the west might be explained by population
density in these regions. It could also be that the dataset simply does not 
include as much information from these areas due to the way the data was 
collected. Information about this is currently not available to
me, so I'll leave that open for further research.

When we investigate the relations the number of people killed and injured for
each incident, we don't find much of a meaningful relationship in general. It
appears that for high numbers of people involved, the numbers are positively
correlated. For small numbers of killings, the number of injured people also
tends to be small (and vice versa).

```{r, killed-injured, echo=FALSE}
ggplot() +
  geom_jitter(data = inc, aes(x = n_killed, y = n_injured), alpha = .2)
```

When we zoom into the bottom-left corner of the previous plot, we find a 
triangular relationship between the number of killings and the number of injured
people. This is backed up by the plot for people involved in an incident; we 
already know that it is very likely for an incident to involve less than ten
people, and very unlikely for an incident to involve more than ten. The
triangular shape in this plot occurs since in this dataset, the number of people
involved in an incident is the sum of people killed and injured in it. 
Interestingly, the solid triangular shape suggests that the probability
distribution of being injured or killed is actually uniform; that is it would be
equally likely to bekilled or injured when being involved in an incident. 
However, this is a global observation and does not take into account the type of
crime. For a gang fight, for example, it could be assumed that every participant
is actively engaging in the crime - if that type of incident would be
the majority, the result (being killed or surviving injured) would be governed 
by chance.

```{r, killed-injured-zoom, echo=FALSE, warning=FALSE}
ggplot() +
  geom_jitter(data = inc, aes(x = n_killed, y = n_injured), alpha = .2) +
  scale_x_continuous(limits=c(0, 10)) +
  scale_y_continuous(limits=c(0, 10))
```

When splitting participant gender statistics into subject and victim groups,
we find a picture that is generally proportional to the overall distribution we
saw before. While the number of male participants is significantly larger
than the number of female or unknown participants, the number of subjects and
suspects is roughly the same as the number of victims.
We do find antiproportional behavior between male and femals groups though:
Where there are more male subjects and suspects than victims, there are
more female victims than subjects and suspects.

```{r, gender-type, echo=FALSE, warning=FALSE}
ggplot() +
  geom_count(data=par, mapping=aes(x=participant_type, y=participant_gender,
                                   shape=participant_gender, color=participant_type)) +
  scale_size_area(max_size=14)
```

When we investigate the relative distribution of
subject/suspect and victim within each gender group, we find a
much clearer picture: Not only there are significantly less women involved
in gun related incidents, more than 72.3% of them are victims.
In contrast, 55.2% of all men involved are subjects or suspects.

```{r, gender-type-bar, echo=FALSE, warning=FALSE}
ggplot(aes(x=participant_type), data=par) +
  geom_bar(aes(y = (..count..)/tapply(..count..,..PANEL..,sum)[..PANEL..]), width = 0.5) +
  facet_grid(rows = vars(participant_gender)) +
  scale_y_continuous(labels=percent)
```

Here's the actual counts and percentages normalized per gender, excluding
unknown gender:

```{r, gender-type-statistics, echo=FALSE}
female.count <- as.integer(count(par[(par$participant_gender == 'Female'), ]))
male.count <- as.integer(count(par[(par$participant_gender == 'Male'), ]))

victims.female.count <- as.integer(count(par[(par$participant_gender == 'Female') & 
                                               (par$participant_type == 'Victim'), ]))
victims.female.percentage <- victims.female.count / female.count

subject.female.count <- as.integer(count(par[(par$participant_gender == 'Female') & 
                                               (par$participant_type == 'Subject-Suspect'), ]))
subject.female.percentage <- subject.female.count / female.count

victims.male.count <- as.integer(count(par[(par$participant_gender == 'Male') & 
                                             (par$participant_type == 'Victim'), ]))
victims.male.percentage <- victims.male.count / male.count

subject.male.count <- as.integer(count(par[(par$participant_gender == 'Male') & 
                                             (par$participant_type == 'Subject-Suspect'), ]))
subject.male.percentage <- subject.male.count / male.count

gender <- as.factor(c('Female', 'Female', 'Male', 'Male'))
type <- as.factor(c('Victim', 'Subject/Suspect', 'Victim', 'Subject/Suspect'))
proportion <- unlist(c(victims.female.percentage, subject.female.percentage, 
                        victims.male.percentage, subject.male.percentage))
counts <- unlist(c(victims.female.count, subject.female.count, 
                        victims.male.count, subject.male.count))
participant_stats <- data.frame('Gender' = gender, 'Type' = type, 
                                'Proportion' = proportion, 'Count' = counts)
participant_stats
```

Before we looked at the age statistics for all participants. Let's now
have a look at the ages of subjects/suspects and victims:

```{r, ages-facet, echo=FALSE}
ggplot(aes(participant_age), data=par[!is.na(par$participant_age), ]) +
    geom_bar(aes(y = (..count..)/sum(..count..)), width = 0.5) +
    scale_y_continuous(labels=percent) +
  facet_grid(rows = vars(participant_type))
```

Interestingly, the distributions don't have a vastly different shape, apart
from showing more younger victims than subjects/suspects, as well as a higher
number in subjects/suspects at ages around 20, compared to victims at that age.
There still seem to be some outliers for the ages, such as victims aged zero
and victims aged 80+:

```{r, ages-summary-subject, echo=FALSE}
par.has_age <- !is.na(par$participant_age)
par.is_subject <- par$participant_type == 'Subject-Suspect'
par.is_victim <- par$participant_type == 'Victim'
summary(par[par.has_age & par.is_subject, "participant_age"])
```

The age statistics for victims do look similar, with the distribution being
somewhat more long-tailed:

```{r, ages-summary-victim, echo=FALSE}
summary(par[par.has_age & par.is_victim, "participant_age"])
```

It turns out that the cases with old subjects/suspects are actually correct,
such as a June 6th, 2015 case where a 98 year old husband apparently killed
his 94 year old wife. The case of the zero-year old subject turns out to be
wrong, however the source material is not available anymore. On the other hand, 
the cases of one year old subjects appear to be correct: These are, for example,
about accidental killings using unsecured guns.

```{r, age-outlier-subjects, echo=FALSE}
sub_extreme_ages <- par[par.has_age & par.is_subject & ((par$participant_age < 2) | (par$participant_age >= 95)), ]
inc_par <- right_join(inc, sub_extreme_ages, by = 'incident_id')
inc_par[, c('incident_id', 'date', 'participant_age', 'incident_url')]
```

When we zoom in on the age group 0-18, we find that there is a slightly higher
number of victims age three compared to any other age below 10. It could be
interesting to investigate the reasons for this.

```{r, ages-facet-young, echo=FALSE}
par_young_ages <- par[par.has_age & par$participant_age < 18, ]
ggplot(aes(participant_age), data=par_young_ages) +
    geom_bar(aes(y = (..count..)/sum(..count..)), width = 0.5) +
    scale_y_continuous(labels=percent) +
  facet_grid(rows = vars(participant_type))
```

## Multivariate plots

Revisiting the previously drawn map, we can also identify incidents where
either a high number of people involved were killed (red marker) or 
injured (yellow).

```{r map-of-incidents-10, echo=FALSE, warning=FALSE}

incident_with_coords_10k <- subset(inc, valid_locations & (inc$n_killed > 10))
incident_with_coords_10i <- subset(inc, valid_locations & (inc$n_injured > 10))

ggplot() + 
  coord_map(xlim = c(-180, -60), ylim = c(10, 70)) +
  geom_polygon(data = map, aes(x = long, y = lat, group = group), fill = 'gray22', alpha = 1) +
  geom_point(data = incident_with_coords_10i, aes(x = longitude, y = latitude), color = 'yellow', alpha = 0.75) +
  geom_point(data = incident_with_coords_10k, aes(x = longitude, y = latitude), color = 'red', alpha = 0.75)
```

To connect the map with the previous age observations, let's have a look
at the locations of incidents by age group.

```{r map-of-incidents-by-age, echo=FALSE, warning=FALSE}
inc_par <- inner_join(inc, par, 'incident_id')

ggplot() + 
  coord_map(xlim = c(-180, -60), ylim = c(10, 70)) +
  geom_polygon(data = map, aes(x = long, y = lat, group = group), fill = 'gray22', alpha = 1) +
  geom_point(data = inc_par, aes(x = longitude, y = latitude), color = 'red', alpha = 0.01) + 
  facet_grid(participant_type ~ participant_age_group_ordered)
```

We find that the general appearance of the map doesn't change much, indicating
that locations of incidents are generally independent of age.

TODO: Distribution of subject age in relation to victim age by gender.
